Real Time Genome Scan Using GPU

### 2019-07-18 paper.tex
modified timing in table with hippocampus data

\author[$\ast$]{Xiaoqi Hu}
\author[$\ast$]{Hyeonju Kim}
\author[$\ast$]{GregoryFarage}
\author[$\dagger$]{PjotrPrins}
\author[$\dagger$]{RobertW.Williams}
\author[$\S$]{KarlW.Broman}
\author[$\ast$]{Saunak Sen}

@@

\subsection{Acceleration Techniques}
We used various techniques to accelerate eQTL scans. 
\subsubsection{Multithreaded CPU operations}
 Used for multi-threaded loops whenever possible, eg. for element-wise operations
\subsubsection{Matrix and Vectorized Operations}
% OpenBLAS for multi-threaded matrix operations. 
Matrix multiplication is a well studied area where the computer scientists try to apply various optimization techniques to make it run faster. 
There are various BLAS libraries available, such as gslBLAS, OpenBLAS, etc. 
The matrix multiplication in OpenBLAS is default multithreaded, therefore does not require extra coding effort to achieve highly efficient result.

@@

\subsubsection{Julia language}
%Programming simplicity of interpreted languages with speed approaching compiled languages; GPU programming and linear algebra straightforward
In our experiment of matrix multiplication, Julia's speed is comparable to C/C++. 
However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative and CuArrays affords much lower programming efforts compared to C/C++.
%TODO: Add reference to CUDAnative and CuArrays
\subsubsection{Heritability grid}
For LMMs, considering a discrete grid of heritabilities enhances parallelization

@@

\begin{table*}[htbp]
	\renewcommand{\familydefault}{\sfdefault}\normalfont
	\centering
	\caption{\bf Time comparison between CPU and GPU}
	\begin{tableminipage}{\textwidth}
		\begin{tabularx}{\textwidth}{XXXX}
			\hline
			\header Method & Precision & CPU only   & CPU \& GPU  \\
			\hline
			\multirow{2}{*}{No Find Max LOD} & Single    & 0.36s & 0.41s \\
			& Double    & 0.55s & 0.80s \\
			\multirow{2}{*}{Find Max LOD}    & Single    &1.54s       &0.056s       \\
			& Double    & 1.77s & 0.079s \\
			\hline

@@

\subsection{Benefit of using GPU}
In parallel computing, Amdahl's law is used to find the theoretical maximum speedup by improving a particular part of a program.  
For example, if a program took ten minutes for a serial processor, and a function that takes nine our of those ten minutes can be parallelized, then, the theoretical speedup, no matter how many processors are used, can not be more than ten times, because the minimum execution time of this program is one minute. 
Therefore, profiling the entire genome scan process is the prerequisit in order to begin the optimization. 
Often, profiling would consider space and time complexity. 
Our primary concern is the time taken by each function, therefore only timing information is considered in our profiling. 
We used Julia's built-in Sampling Profiler (%TODO reference here.)
to find our target function for GPU because it is less intrusive compared to other profiling methods.

@@

To cope with this limitaion of GPU, we further developed \textit{Find Max LOD} method, because main interest of genome scan is to find the maximum LOD score of every phenotype. 
This step is highly parallelizable and can utilize GPU's massive cores, at the same time, reduce the amount of data needed to be transfered back to host. 
Table \ref{tab:speedup-table} shows that \textit{Find Max LOD} method reduces CPU\&GPU execution time down to 0.079 second when using double precision. 
Compared to CPU only, which took 1.77 seconds, this method gains 22 times speedup.

>>>>

\author[$\ast$]{Xiaoqi Hu}
\author[$\ast$]{Hyeonju Kim}
\author[$\ast$]{Gregory Farage}
\author[$\dagger$]{Pjotr Prins}
\author[$\dagger$]{Robert W. Williams}
\author[$\S$]{Karl W. Broman}
\author[$\ast$]{Saunak Sen}

@@

\subsection{Acceleration Techniques}
We used various techniques to accelerate eQTL scans. 
\subsubsection{Multithreaded CPU operations}
% Used for multi-threaded loops whenever possible, eg. for element-wise operations.

\subsubsection{Matrix and Vectorized Operations}

% OpenBLAS for multi-threaded matrix operations. 
Since our algorithm largely depends on matrix operations, it is natual to find the fastest way to achieve best result. 
There are various matrix libraries available, such as gslBLAS and OpenBLAS. 
They target different hardware or use various techniques to get highly optimal results. 
The matrix multiplication in OpenBLAS is default multithreaded, therefore does not require extra coding effort to make the CPU version of matrix multiplication run in parallel.

Since matrix multiplication and element-wise operation are algorithmically free of data and function dependency, it is amendable to GPU's parallel computing power. 
We run matrix multiplication with different shapes matrices, and compared the running time of CPU and GPU. 
The result is shown in Figure \ref{GPUCPUShape}

@@

\subsubsection{Julia language}
%Programming simplicity of interpreted languages with speed approaching compiled languages; GPU programming and linear algebra straightforward
In our experiment of matrix multiplication, Julia's speed is comparable to C/C++. 
However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative and CuArrays affords much lower programming efforts compared to C/C++.
%TODO: Add reference to CUDAnative and CuArrays

@@

\begin{table*}[htbp]
	\renewcommand{\familydefault}{\sfdefault}\normalfont
	\centering
	\caption{\bf Time comparison between CPU and GPU}
	\begin{tableminipage}{\textwidth}
		\begin{tabularx}{\textwidth}{XXXXX}
			\hline
			\header Dataset & Method & Precision & CPU only   & CPU \& GPU  \\
			\hline
			\multirow{4}{*}{Spleen Data}                                                & \multirow{2}{*}{No Find Max} & Single    & 0.36s  & 0.41s \\
			&                              & Double    & 0.55s  & 0.80s \\
			& \multirow{2}{*}{Find Max}    & Single    & 0.46s  & 0.06s \\
			&                              & Double    & 0.71s  & 0.08s \\
			\hline
			\multirow{2}{*}{\begin{tabular}[c]{@{}c@{}}Hippocampus\\ Data\end{tabular}} & \multirow{2}{*}{Find Max}    & Single    & 17.64s & 2.26s \\
			&                              & Double    & 26.77s & 3.51s \\
			\hline

@@

\subsection{Benefit of using GPU}
In parallel computing, Amdahl's law is used to find the theoretical maximum speedup by improving a particular part of a program.  
For example, if a program took ten minutes for a serial processor, and a function that takes nine our of those ten minutes can be parallelized, then, the theoretical speedup, no matter how many processors are used, can not be more than ten times, because the minimum execution time of this program is one minute. 
Therefore, profiling the entire genome scan process is the prerequisit in order to begin the optimization. 
Often, profiling would consider space and time complexity. 
Our primary concern is the time taken by each function, therefore only timing information is considered in our profiling. 
We used Julia's built-in Sampling Profiler %TODO reference here.)
to find our target function for GPU because it is less intrusive compared to other profiling methods.

@@

To cope with this limitaion of GPU, we further developed \textit{Find Max LOD} method, because main interest of genome scan is to find the maximum LOD score of every phenotype. 
This step is highly parallelizable and can utilize GPU's massive cores, at the same time, reduce the amount of data needed to be transfered back to host. 
Table \ref{tab:speedup-table} shows that \textit{Find Max LOD} method reduces CPU\&GPU execution time down to 0.079 second when using double precision. 
Compared to CPU only, which took 1.77 seconds, this method provides 22 times speedup.
